Glacial flows, simulated faster

### 2023-06-05 talk-iu/slides.tex
good edits ... it is alreaady a goot talk

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is a \emph{very viscous, incompressible, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more soon \dots
    \end{itemize}
\item glaciers lie on \emph{topography}
    \begin{itemize}
    \item[$\circ$] except sometimes they float on water (floating tongue or ice shelf)
    \end{itemize}
\item a glacier's geometry (\emph{free surface}), and its velocity, \emph{evolve in contact with the climate}:
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] sub-shelf melt (when floating)
    \item[$\circ$] calving (into ocean)
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{../images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}
}

@@

\begin{frame}{glaciers}

@@

\begin{itemize}
\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width
\end{itemize}

\bigskip
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{center}
\includegraphics<1>[height=0.69\textheight]{../images/ant-pittard2021.png}
\only<1>{\par {\scriptsize Antarctic ice sheet} \hfill {\tiny (Pittard et al 2021)}}
\only<2>{\vspace{17mm}}
\includegraphics<2>[width=\textwidth]{../images/ant-schoofhewitt2013.png}
\only<2>{\vspace{4mm}\par {\scriptsize note smooth surface and rough bed \dots and vertical exaggeration} \hfill {\tiny (Schoof \& Hewitt 2013)}}
\includegraphics<3>[height=0.69\textheight]{../images/alps-seguinot2018.png}
\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum} \hfill {\tiny (Seguinot et al 2018)}}
\includegraphics<4>[height=0.69\textheight]{../images/moraines-indiana.jpg}
\only<4>{\par {\scriptsize moraines in Illinois, Indiana, Ohio} \hfill {\tiny (Larsen 1986 and other sources)}}
\includegraphics<5>[height=0.69\textheight]{../images/laurentide-margold2018.png}
\only<5>{\par {\scriptsize Laurentide ice sheet} \hfill {\tiny (Margold, Stokes, Clark 2018)}}
\end{center}
\end{minipage}
\end{frame}

\begin{frame}{an ice sheet is \emph{not} sea ice!}
\begin{center}
\includegraphics[height=0.8\textheight]{../images/not-sea-ice.png}
\end{center}
\end{frame}

@@

\begin{frame}{simplifications}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming modeling, I will ignore parts of glacier physics in the remainder of my talk

\medskip
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (e.g.~calving)
    \item[$\circ$] solid earth deformation
    \end{itemize}

\medskip
\item {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, which includes these and other processes}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{../images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

\section{the basic glacier model}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
a \alert{glacier} or \alert{ice sheet model} is a \underline{map}

which evolves a glacier in a climate
\end{definition} 
\item at least two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
\text{precipitation minus} \\
\text{melt \& runoff}
\end{pmatrix}$$

@@

\smallskip
\item {\small \alert{map}: \, $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$ }
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{../images/map-velocity.png}

@@

\begin{frame}{basic model: notation}

\begin{columns}
\begin{column}{0.55\textwidth}
\begin{itemize}
\item data $a(t,x,y)$, $b(x,y)$ are defined on a \alert{fixed domain}:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
\item<2-> solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] also a fixed domain,
    \item[$\circ$] but $s=b$ where there is no ice
    \end{itemize}
\item<3-> $s(t,x,y)$ determines the \alert{icy domain} $\Lambda(t) \subset \RR^3$:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$
\item<4> the solution velocity $\bu(t,x,y,z)$ is defined on $\Lambda(t)$

@@

\begin{frame}{basic model: conservation}

\begin{itemize}
\item ice sheet evolution should conserve physical quantities:
    \begin{itemize}
    \item[$\circ$] mass
    \item[$\circ$] momentum
    \item[$\circ$] energy \hfill $\leftarrow$ \emph{ignored for simplicity in this talk}
    \end{itemize}

\medskip
\item conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] in the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t) \\
    \intertext{\item[$\circ$] and on the ice surfaces:}
\text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t) \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t)
\end{align*}

        \begin{itemize}
        \item[$\vartriangleright$] $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
        \item[$\vartriangleright$] $\bn_s = \left<-\grad s,1\right>$ is upward surface normal
        \end{itemize}
    \end{itemize}

@@

\begin{frame}{free boundary problem}

\begin{itemize}
\item ice sheet evolution is a \alert{free-boundary} problem for conserved quantities
\item specifically, the surface kinematic equation (SKE)
  $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s = a$$
applies \emph{only} on the ice upper surface $\Gamma_s(t)$
\item in the remainder of the (fixed) domain $\Omega\subset \RR^2$, \alert{complementarity} holds:
  $$s=b \quad \text{and} \quad a \le 0$$

\bigskip
\item {\footnotesize for more on this perspective see Bueler (2021)}
\end{itemize}
\end{frame}

\begin{frame}{basic model: strong form = NCP coupled to Stokes}

@@

    \begin{itemize}
    \item[$\circ$] note: $\bu|_s=0$ where no ice
    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu(D\bu) = \Gamma |D\bu|^{\pp-2}$
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{the basic model is a DAE system}

\begin{itemize}
\item for this slide, forget complementarity and boundary conditions
\item result: SKE coupled to Stokes
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 \\
- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
\nabla \cdot \bu &= 0
\end{align*}

\item only the first of these 5 equations has a time derivative
    \begin{itemize}
    \item[$\circ$] because ice is very viscous and incompressible
    \end{itemize}
\item<2> this time-dependent problem is a \alert{differential algebraic equation} (DAE), an extremely stiff system:
\begin{align*}
\dot x &= f(x,y) \\
     0 &= g(x,y)
\end{align*}

    \begin{itemize}
    \item<2>[$\circ$] but in $\infty$ dimensions \dots and actually subject to complementarity
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{basic model: current research}

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
    \begin{itemize}
    \item[$\circ$] though most researchers would agree NCP-coupled-to-Stokes is indeed the intended model \dots
    \end{itemize}

@@

\begin{frame}{basic model: current \emph{thinking}}

@@

\medskip
\item \dots and ice sheet geometry evolution is addressed with minimal awareness of complementarity

\medskip
\item the NCP-coupled-to-Stokes basic model is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
    \begin{itemize}
    \item[$\circ$] because it is too slow
    \item[$\circ$] \alert{can we make it fast enough to use?}
    \end{itemize}
\end{itemize}
\end{frame}

\section{numerics: time-stepping and solver scaling}

\begin{frame}{ice sheet models: the mass-continuity equation view}

\begin{itemize}
\item thickness transport form helps for evolution or stability questions
\item define:
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}

    \begin{itemize}
    \item[$\circ$] note $s$ and $H$ are equivalent variables for modeling ice geometry
    \end{itemize}
\item the \alert{mass continuity equation} for thickness follows from SKE and incompressibility:
   $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$

\smallskip
\item<2> \emph{question:} is this really an advection equation?
\item<2>[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H$$
\item<2> observation: the NCP-coupled-to-Stokes system has no characteristic curves
\end{itemize}
\end{frame}

@@

\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item the diffusion schema is literal in the shallow ice approximation
    \begin{itemize}
    \item[$\circ$] more on this momentarily
    \end{itemize}
\item the fact that ice flows downhill has \emph{time-stepping stability} consequences for the basic model
    \begin{itemize}
    \item[$\circ$] regardless of your above schema preference!
    \end{itemize}
\item note both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
\end{itemize}
\end{frame}

\begin{frame}{shallow ice approximation NCP}

\begin{itemize}
\item as mentioned, the simplest of glacier flow approximations is the ``lubrication'' approximation: \alert{shallow ice approximation} (SIA)
\item SIA version of the NCP:
{\small
  $$s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0$$
}
the surface motion contribution ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$ has a formula:
  $${\color{FireBrick} \Phi(s)} = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)$$

\medskip
\item $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
    \begin{itemize}
    \item[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
    \item[$\circ$] \emph{local} in surface and bed topography
    \item[$\circ$] well-posedness for this problem as weak form \alert{variational inequality} (Calvo et al 2002, Jouvet \& Bueler 2012, Piersanti \& Temam 2022)
    \end{itemize}
\end{itemize}
\end{frame}

\begin{itemize}
\item from now on, let us avoid shallowness approximations
\item then the basic ice sheet model (NCP coupled to Stokes) problem has a \alert{non-local} surface velocity function ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$
{\small
\begin{equation*}
s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0
\end{equation*}
}
\item \emph{figure:} the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the SIA velocity responds only underneath the surface perturbation}
\end{itemize}

@@

\begin{minipage}[t][60cm][t]{\textwidth}
\begin{itemize}
\only<1>{
\item let us recall some traditional numerical analysis
}
\only<2>{
\item \alert{explicit} time stepping is common for \alert{advections}
\item for example, forward Euler using spacing $h$ and time step $\Delta t$:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{\bq_{j+1/2}^\ell - \bq_{j-1/2}^\ell}{h} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] need good approximations of flux $\bq=\bU H$: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \item[$\circ$] conditionally stable, with CFL maximum time step
        $$\Delta t \le \frac{h}{\max |\bU|} = O(h)$$
    \end{itemize}
}
\only<3>{
\item \alert{explicit} time stepping for \alert{diffusions} is best avoided
\item for example, forward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell} + s_{j}^{\ell}) - D_{j-\frac{1}{2}} (s_{j}^{\ell} + s_{j-1}^{\ell})}{h^2} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] conditionally stable, with maximum time step
        $$\Delta t \le \frac{h^2}{\max D} = O(h^2)$$
    \end{itemize}
}
\only<4>{
\item \alert{implicit} time stepping for \alert{diffusions} is often recommended
\item for example, backward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell+1} + s_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (s_{j}^{\ell+1} + s_{j-1}^{\ell+1})}{h^2} = a_j^\ell$$

@@

\begin{itemize}
\item current-technology large-scale models use explicit time stepping
    \begin{itemize}
    \item[$\circ$] this is embarrassing: the mathematical problem is a DAE
    \item[$\circ$] the accuracy/performance/usability consequences of the suppressed free-boundary/DAE/diffusive character are hard to sweep under the rug
    \end{itemize}
\item yet most researchers believe the advection schema
    \begin{itemize}
    \item[$\circ$] time step is determined by CFL using coupled solution velocity $\bU$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \alert{implicit time-stepping} is appropriate for DAE problems
\item a sequence of NCP-coupled-to-Stokes free-boundary problems must be solved at each time step
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item a non-shallow model solves a Stokes problem at each step:
\begin{align*}
- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu(D\bu) D\bu - pI\right) \bn_s &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item this is the \alert{stress balance} (conservation of momentum) problem which determines velocity $\bu$ and pressure $p$
\item how fast is the numerical solution process?
    \begin{itemize}
    \item[$\circ$] how do solution algorithms \alert{scale} with increasing spatial resolution?
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{summary: PDE solver algorithmic scaling}

>>>>

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is a \emph{very viscous, incompressible, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more soon \dots
    \end{itemize}
\item glaciers lie on \emph{topography}
    \begin{itemize}
    \item[$\circ$] except sometimes they float on water (floating tongue or ice shelf)
    \end{itemize}
\item a glacier's geometry (\emph{free surface}), and its velocity, \emph{evolve in contact with the climate}:
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] sub-shelf melt (when floating)
    \item[$\circ$] calving (into ocean)
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{../images/map-velocity.png}
\end{column}
\end{columns}
\end{frame}
}

@@

\begin{frame}{pictures of glaciers}

@@

\begin{itemize}
\item \emph{def.} \, \alert{ice sheet} $=$ a large glacier with small thickness$/$width ratio
\end{itemize}

\bigskip
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{center}
\includegraphics<1>[height=0.69\textheight]{../images/ant-pittard2021.png}
\only<1>{\par {\scriptsize Antarctic ice sheet} \hfill {\tiny (Pittard et al 2021)}}
\only<2>{\vspace{17mm}}
\includegraphics<2>[width=\textwidth]{../images/ant-schoofhewitt2013.png}
\only<2>{\vspace{4mm}\par {\scriptsize note smooth surface and rough bed \dots and vertical exaggeration} \hfill {\tiny (Schoof \& Hewitt 2013)}}
\includegraphics<3>[height=0.69\textheight]{../images/alps-seguinot2018.png}
\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum} \hfill {\tiny (Seguinot et al 2018)}}
\includegraphics<4>[height=0.69\textheight]{../images/laurentide-margold2018.png}
\only<4>{\par {\scriptsize Laurentide ice sheet, $\approx$ 22,000 years ago} \hfill {\tiny (Margold, Stokes, Clark 2018)}}
\includegraphics<5>[height=0.69\textheight]{../images/moraines-indiana.jpg}
\only<5>{\par {\scriptsize moraines in Illinois, Indiana, Ohio} \hfill {\tiny (Larsen 1986 and other sources)}}
\end{center}
\end{minipage}
\end{frame}

\begin{frame}{finally, an ice sheet is \emph{not} sea ice!}
\begin{center}
\includegraphics[height=0.8\textheight]{../images/not-sea-ice.png}
\end{center}
\end{frame}

\section{the basic mathematical model for glaciers}

@@

\begin{frame}{modeling simplifications}

\begin{itemize}
\item for simplicity/clarity of the upcoming model, I will \alert{ignore} these aspects of glacier physics in my talk:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (e.g.~calving)
    \item[$\circ$] solid earth deformation
    \end{itemize}

\medskip
\item all are important for doing science!
\item UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, includes these and other processes
\end{itemize}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
a \alert{glacier model} is a \underline{map}

which evolves a glacier in a climate
\end{definition} 
\item at least two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
\text{snowfall minus} \\
\text{melt \& runoff}
\end{pmatrix}$$

@@

\smallskip
\item<2> {\small \underline{map}: \, $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$ }
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{../images/map-velocity.png}

@@

\begin{frame}{the basic glacier model: notation}

\begin{columns}
\begin{column}{0.55\textwidth}
\begin{itemize}
\item data $a(t,x,y)$, $b(x,y)$ are defined on a \alert{fixed domain}:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
\item<2-> solution \alert{surface elevation} $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] also a fixed domain,
    \item[$\circ$] but $s=b$ where there is no ice
    \end{itemize}
\item<3-> $s(t,x,y)$ determines the \alert{icy domain} $\Lambda(t) \subset \RR^3$:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$
\item<3-> the solution \alert{velocity} $\bu(t,x,y,z)$ is defined on $\Lambda(t)$

@@

\begin{frame}{the basic glacier model: conservation}

\begin{itemize}
\item glacier evolution is merely physics \dots so it \alert{conserves}
    \begin{itemize}
    \item[$\circ$] mass
    \item[$\circ$] momentum
    \item[$\circ$] energy \hfill $\leftarrow$ \emph{ignored for simplicity in this talk}
    \end{itemize}

\medskip
\item<2> conservation of mass happens
    \begin{itemize}
    \item[$\circ$] within the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t) \\
    \intertext{\item[$\circ$] on the surfaces $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$:}
\text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t) \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t)
\end{align*}

        \begin{itemize}
        \item[$\vartriangleright$] $\Gamma_s(t)$ is upper surface of the ice
        \item[$\vartriangleright$] $\Gamma_b(t)$ is base of the ice
        \item[$\vartriangleright$] $\bn_s = \left<-\grad s,1\right>$ is upward surface normal
        \end{itemize}
    \end{itemize}

@@

\begin{frame}{the free boundary problem for a fluid layer}

\begin{itemize}
\item glacier evolution is a \alert{free-boundary} problem
\item specifically, the surface kinematic equation (SKE)
  $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s = a$$
applies \emph{only} on the ice upper surface $\Gamma_s(t)$
\item in the remainder of the (fixed) domain $\Omega\subset \RR^2$, \alert{complementarity} holds:
  $$s=b \quad \text{and} \quad a \le 0$$

\bigskip
\item {\footnotesize for more on this perspective see Bueler (2021), \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math}
\end{itemize}
\end{frame}

\begin{frame}{the basic glacier model strong form: \, NCP coupled to Stokes}

@@

    \begin{itemize}
    \item[$\circ$] note: $\bu|_s=0$ where no ice
    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu(D\bu) = \Gamma |D\bu|^{\pp-2}$, \,$p \approx 4$
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{the basic glacier model is a DAE system}

\begin{itemize}
\item for this slide, forget complementarity and boundary conditions to get simplified model ``SKE coupled to Stokes'':
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 \\
- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
\nabla \cdot \bu &= 0
\end{align*}

\item only the first of these 5 equations has a time derivative
    \begin{itemize}
    \item[$\circ$] recall: ice is very viscous and incompressible
    \end{itemize}
\item<2> this time-dependent problem is a \alert{differential algebraic equation} (DAE), an extremely stiff system:
\begin{align*}
\dot x &= f(x,y) \\
     0 &= g(x,y)
\end{align*}

    \begin{itemize}
    \item<2>[$\circ$] in $\infty$ dimensions, of course,
    \item<2> and also subject to complementarity
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{the basic glacier model: current research}

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
    \begin{itemize}
    \item[$\circ$] though most researchers would agree NCP-coupled-to-Stokes is indeed the intended model!
    \end{itemize}

@@

\begin{frame}{the basic glacier model: current numerical \emph{thinking}}

@@

\medskip
\item \dots and ice sheet geometry evolution is often addressed with minimal awareness of complementarity

\medskip
\item<2> the NCP-coupled-to-Stokes basic model is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
    \begin{itemize}
    \item[$\circ$] because it is too slow
    \item[$\circ$] \alert{can we make it fast enough to use?} \hfill \begin{minipage}{0.3\textwidth}{\footnotesize \color{FireBrick}{$\leftarrow$ \emph{what I am working on}}} \end{minipage}
    \end{itemize}
\end{itemize}
\end{frame}

\section{numerics: time-stepping}

\begin{frame}{the mass-continuity equation view}

\begin{itemize}
\item ``thickness transport form'' helps for evolution or stability questions
\item define:
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}

    \begin{itemize}
    \item[$\circ$] note $s$ and $H$ are equivalent variables for modeling ice geometry
    \end{itemize}
\item the \alert{mass continuity equation} for thickness, an apparent \alert{advection equation}, follows from the SKE and incompressibility:
   $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$

\smallskip
\item<2> \emph{question:} is this really an advection equation?
\item<2>[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H$$
\item<2> in any case, the NCP-coupled-to-Stokes system \emph{has no characteristic curves}
\end{itemize}
\end{frame}

@@

\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item both forms are nonlinear: \, $\bU=\bU(H,\grad s), \, D=D(H,\grad s)$
\item the glacier modeling literature is confusing!
\item the diffusion schema is literal in the shallow ice approximation
    \begin{itemize}
    \item[$\circ$] more on this momentarily
    \end{itemize}
\item regardless of your schema preference, the fact that ice flows downhill has \emph{time-stepping stability} consequences!
\item \dots so let us recall some traditional numerical analysis
\end{itemize}
\end{frame}

@@

\begin{minipage}[t][60cm][t]{\textwidth}
\begin{itemize}
\only<1>{
\item \alert{explicit} time stepping is common for \alert{advections}
\item for example, forward Euler using spacing $h$ and time step $\Delta t$:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{\bq_{j+1/2}^\ell - \bq_{j-1/2}^\ell}{h} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] need good approximations of flux $\bq=\bU H$: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \item[$\circ$] conditionally stable, with CFL maximum time step
        $$\Delta t \le \frac{h}{\max |\bU|} = O(h)$$
    \end{itemize}
}
\only<2>{
\item \alert{explicit} time stepping for \alert{diffusions} is best avoided
\item for example, forward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell} + s_{j}^{\ell}) - D_{j-\frac{1}{2}} (s_{j}^{\ell} + s_{j-1}^{\ell})}{h^2} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] conditionally stable, with maximum time step
        $$\Delta t \le \frac{h^2}{\max D} = O(h^2)$$
    \end{itemize}
}
\only<3>{
\item \alert{implicit} time stepping for \alert{diffusions} is often recommended
\item for example, backward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell+1} + s_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (s_{j}^{\ell+1} + s_{j-1}^{\ell+1})}{h^2} = a_j^\ell$$

@@

\begin{itemize}
\item current-technology, large-scale numerical models, including \href{https://www.pism.io/}{PISM}, use explicit time stepping
    \begin{itemize}
    \item[$\circ$] this is embarrassing: the mathematical problem is a DAE
    \end{itemize}
\item many researchers ``believe'' the advection schema
    \begin{itemize}
    \item[$\circ$] time step is supposed to be determined by CFL using the coupled solution velocity $\bU$
    \end{itemize}
\item the accuracy/performance/usability consequences of the suppressed DAE/diffusive character are hard to sweep under the rug
\item the whole situation is a cry for mathematical clarity!
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \alert{implicit time-stepping} is appropriate for DAE problems
\item future models will solve a sequence of NCP-coupled-to-Stokes free-boundary problems at each time step
\end{itemize}
\end{frame}

\section{numerics: Stokes models}

\begin{frame}{shallow ice approximation \quad $\leftarrow$ \emph{not Stokes}}

\begin{itemize}
\item the simplest of glacier flow approximations is the ``lubrication'' approximation: \alert{shallow ice approximation} (SIA)
\item SIA version of the NCP:
{\small
  $$s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0$$
}
the surface motion contribution ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$ has a formula:
  $${\color{FireBrick} \Phi(s)} = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)$$

\medskip
\item ${\color{FireBrick} \Phi(s)}$ is a \alert{doubly-nonlinear differential operator}
    \begin{itemize}
    \item[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
    \item[$\circ$] but \emph{local} in surface and bed topography, which Stokes is not
    \item[$\circ$] well-posedness holds for the weak form $=$ \alert{variational inequality} (Calvo et al 2002, Jouvet \& Bueler 2012, Piersanti \& Temam 2022)
    \end{itemize}
\end{itemize}
\end{frame}

\begin{itemize}
\item however, from now on, let us avoid shallowness approximations
\item the basic glacier model (NCP coupled to Stokes) problem has a \alert{non-local} surface velocity function ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$
{\small
\begin{equation*}
s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0
\end{equation*}
}
\item the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the SIA velocity responds only underneath the surface perturbation}
\end{itemize}

@@

\begin{itemize}
\item a non-shallow model solves a Stokes problem at each step:
\begin{align*}
- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu(D\bu) D\bu - pI\right) \bn_s &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item this is the \alert{stress balance} (conservation of momentum) problem which determines velocity $\bu$ and pressure $p$
\item<2> how fast is the numerical solution process?
    \begin{itemize}
    \item[$\circ$] how do solution algorithms \alert{scale} with increasing spatial resolution?
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{one slide course: numerical PDE solver algorithmic scaling}
